---
title: "__A Comprehensive Analysis of Sleeping Patterns and Disorders Among US Adults, 2017–2020__"
subtitle: "Kruthika Gaddam, Likith Sai Gowni, Kavya Gustala, Mounisha Madala, Sowmya Papagari"
author: "INFO-B 518 Applied Statistical Methods for Biomedical Informatics"
date: "Indiana University-Purdue University, Indianapolis, USA"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
```
# 1 Project Scope
## 1.1 Introduction
Sleep, like food, water, and oxygen, is essential to human life. The timing of 
sleep is influenced by biological factors, as evidenced by the human tendency 
to sleep at night. However, many factors, including education, occupation, 
social activities, insomnia, and various sleep and medical disorders, 
can result in individuals sleeping at varying times (Grandner, 2017, p. 1). 
The internet has significantly changed people's personal and professional lives,
directly or indirectly influencing the sleep cycle, such as accessing social media 
in bed or browsing late at night (Grandner, 2017, p. 2).


   Circadian rhythms and sleep habits are currently being altered in individuals.
Due to various causes, they rely excessively on weekdays and weekends, negatively 
affecting metabolic function and increasing the risk of many diseases. Because of 
the differing sleep cycles on weekends and weekdays, a substantial percentage of 
sleep debt is compensated on weekends. Social jetlag is the difference between weekends 
and weekdays, as well as between social time and circadian rhythm (Roenneberg et 
al., 2012, p. 939). For instance, the prevalence of adults who sleep less than 6 hours may have 
increased over time in the United States, according to the minimal statistics available 
(Ford et al., 2015, p. 829). Finding out the usual time people go to sleep and wake 
up during the week and on weekends can help healthcare professionals understand sleep
patterns and how they affect health. This can help promote healthy sleep habits and 
improve the health of people in the US.

## 1.2 Aim
The aim of this study is to find out how the people of the United States sleep and 
to evaluate the following things:

1. This study will look at how long people sleep on average, when they wake up on 
average, and how long people sleep on average on weekdays and weekends. 
2. Determine the amount of sleep debt and social jetlag.
3. This study aims to investigate whether individuals who maintain a consistent 
sleep pattern throughout the week, including weekends, experience reduced daytime sleepiness.
4. This study aims to establish a correlation between sleep duration during weekdays 
and weekends and various factors such as snoring and snorting.

## 1.3 Purpose
This project aims to understand the US population's sleeping habits and sleep disturbances. 
With further research, these findings could be used to assist healthcare professionals 
understand sleep patterns and their effects on promoting healthy sleep habits and 
improving the health of US citizens. The results of this study can be used as a basis to investigate and improve the sleep habits of the US population.

## 1.4 Hypotheses
1. Determining if the average sleep-wake time is different on weekends and weekdays.\
 __Null hypothesis__: The average sleep-wake time is not different on weekends and weekdays.\
__Alternate Hypothesis__: The average sleep and wake times differ on weekdays and weekends.\
2. Determining if individuals who have regular sleep cycle is associated with lower 
daytime sleepiness.\
 __Null Hypothesis__: Individuals who have regular sleep cycle is not associated 
with lower daytime sleepiness.\
  __Alternate Hypothesis__: Individuals who have regular sleep cycle is associated 
with lower daytime sleepiness.\
3. Determining if sleep duration on weekdays and weekends is associated with snoring,
snorting, trouble sleeping, or feeling over-sleepy during the day.\
 __Null Hypothesis__: Sleep duration on weekdays and weekends has no association 
with snoring, snorting, difficulty sleeping, or feeling sleepy during the day.\
  __Alternate Hypothesis__: Sleep duration on weekdays and weekends has association 
with snoring, snorting, difficulty sleeping, or feeling sleepy during the day.
  
# 2 Data Exploration
## 2.1 Data collection and extraction
The National Health and Nutrition Examination Survey (NHANES) is an annual survey 
that gathers information on a wide range of health and nutritional issues from a nationally representative sample of the population in the United States of America. The Sleep Disorders 
Dataset (2017-20) comprises 10,196 rows and 11 columns. This data collection contains 
questions derived from the Munich Chrono Type Questionnaire about sleep patterns 
and problems, such as regular sleep times on weekdays and weekends, Sleep hours, Snorting, difficulty sleeping, and feeling drowsy during the day. Participants were surveyed 
at home using the Computer-Assisted Personal Interview (CAPI) technology. The precision and 
comprehensiveness of the data ensured quality assurance.
[Data set Link](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_SLQ.XPT)

```{r message=TRUE, warning=FALSE, include=FALSE, results='hide'}
library(readr)
library(knitr)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
#Loading the csv to import the data
sleepdata = read_csv("C:\\Users\\DELL\\Downloads\\stats_squad_sleepdisorders.csv", show_col_types = FALSE)
kable(head(sleepdata), caption =" Sleep Disorders Dataset")
```

The data cleaning and data analysis were done using RStudio. There were 0.39% missing 
values in the dataset. According to the NHANES, missing values should only be imputed 
if they make up more than 10% of the total observations. In this case, less than 10% 
of the observations were missing values, so they were not imputed. In this project, the observations with missing data were omitted as they were in important variables like
sleep time, wake time, and sleep disturbances, which would have been useless if they
had been kept with the missing data and considered only the complete data for the important variables. The column headers were modified to enhance the understanding of the variables.


## 2.2 Variables
Variables include a Sequence number, weekdays sleep time, weekdays waketime, 
weekdays sleep time, weekends sleep time, weekends wake time, weekends sleep duration, 
Snoring frequency, Snorting, Trouble sleeping, and Excessive sleep during 
the day which are coded as SLQ and SLD. Sequence number is discrete, 
snoring, Snorting and oversleepy during the day are ordinal variables, Trouble 
sleeping is a binary variable, Sleep time, wake time on weekends and week days are 
in the time format and are continuous. Sleep duration on weekends and weekdays are 
continuous variables. In addition other variables like Social jetlag (mid-point d
ifference between sleep on weekdays and weekends) and sleep debt (the difference between 
weekend sleep duration and weekly average sleep duration) will be determined 
(Di et al., 2022, p. 2).

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, results='asis'}

#omitting missing values
sleepdf= na.omit(sleepdata)

#Renaming the column names
colnames(sleepdf)= c('SEQN', 'WD_sleeptime', 'WD_waketime', 'WD_sleephours', 'WE_sleeptime', 'WE_waketime', 'WE_sleephours', 'Snoring', 'Snorting', 'Trouble_sleep', 'Oversleepy')

# Converting sleep and wake times to time format
library(dplyr)
sleepdf= sleepdf %>%
  mutate(across(contains("sleeptime") | contains("waketime"), parse_time))

kable(head(sleepdf), caption =" Cleaned Sleep Disorders Dataset", align = "c",  booktabs = TRUE)


# Calculating sleep debt
sleepdf= sleepdf %>%
  mutate(sleep_debt = WE_sleephours - WD_sleephours)

#Calculating social jetlag
sleepdf = sleepdf %>%
  mutate(mid_sleep_weekdays = WD_sleeptime/3600 + WD_sleephours / 2,
         mid_sleep_weekends = WE_sleeptime/3600 + WE_sleephours / 2,
         social_jetlag = abs(mid_sleep_weekends - mid_sleep_weekdays))

```
## 2.3 Data Distribution and Summary

The summary statistics in Table 2 provide an overview of sleeping patterns. Important 
variables include the duration of sleep on weekdays `WD_sleephours` and weekends 
`WE_sleephours`. The minimum weekday and weekend sleep duration is 3 hours, and the 
maximum is 14 hours. The average weekday sleep duration is 7.65 hours, with a median 
of 7.5 hours. The interquartile range (IQR) is between 7 and 8.5 hours, representing 
the middle 50 percent of the data. The average weekend sleep duration is 8.36 hours, 
with the median being 8 hours as shown in Figure 2. The IQR ranges between 7 and 9.5 hours, representing the median 50% of the data set. The `Snoring` and `Snorting` variables 
ranges from 0 (Never) to 9 (Don't know), with a median of 1 (Rarely) and a median of 0 (Never) 
simultaneously and they both are right skewed (See Appendix). The median value of the trouble sleeping variable `trouble_sleep` is (No), ranging from 1 (Yes) to 9 (Don't know). Over-sleepy during the day variable `Oversleepy` ranges from 0 (Never) to 9 (Don't know), with a mean of 2 
(Sometimes) and is rightly skewed. The average `sleep_debt` is 0.50 hours. Absolute
sleep debt was calculated on weekends and weekdays by taking 8 hours as the recommended 
sleep duration as shown in Figure 2.

```{r, echo=FALSE, results='asis'}

# summary statistics 
vari = c("WD_sleephours", "WE_sleephours", "Snoring", "Snorting", "Trouble_sleep", "Oversleepy", "sleep_debt")
summary_stats <- sapply(sleepdf[vari], summary)
kable(head(summary_stats), caption =" Descriptive Statistics", align = "c",  booktabs = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, results='asis'}
#MEAN SLEEP HOURS
#Mean sleephours on weekdays
weekday_mean = mean(sleepdf$WD_sleephours)
weekday_sd = sd(sleepdf$WD_sleephours)
weekday_n = length(sleepdf$WD_sleephours)

# Calculating the standard error
weekday_se = weekday_sd / sqrt(weekday_n)

# Calculating the 95% confidence interval using the t-distribution
alpha = 0.05
t_critical = qt(1 - alpha/2, df = weekday_n - 1)
weekday_conf_int = c(weekday_mean - t_critical * weekday_se, weekday_mean + t_critical * weekday_se)

weekday_mean
weekday_conf_int


#Mean sleephours on weekends
weekend_mean = mean(sleepdf$WE_sleephours)
weekend_sd = sd(sleepdf$WE_sleephours)
weekend_n = length(sleepdf$WE_sleephours)

# Calculating the standard error
weekend_se = weekend_sd / sqrt(weekend_n)

# Calculating the 95% confidence interval using the t-distribution
alpha = 0.05
t_critical = qt(1 - alpha/2, df = weekend_n - 1)
weekend_conf_int = c(weekend_mean - t_critical * weekend_se, weekend_mean +
                       t_critical * weekend_se)

weekend_mean
weekend_conf_int

#Difference between the sleep duration on weekdays and weekends
Sleepdifference = abs(weekday_mean- weekend_mean)
Sleepdifference
```


```{r fig1, fig.width = 6, fig.asp = .62, echo=FALSE,  results='asis',fig.cap="Distribution of sleep hours" }

## Creating the histograms
# Set up the plotting layout
par(mfrow = c(1, 2))
hist(sleepdf$WD_sleephours, main = "Weekday Sleep Duration", xlab = "Sleep Duration (hours)", ylab = "Frequency", col = "lightcoral", breaks = 20)
hist(sleepdf$WE_sleephours, main = "Weekend Sleep Duration", xlab = "Sleep Duration (hours)", ylab = "Frequency", col = "steelblue", breaks = 20)
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, results='asis'}
#MEAN WAKE TIME
#Mean waketime on weekdays

# Function to convert time objects to seconds
time_to_seconds = function(time) {as.numeric(time, units = "secs")}

# Function to convert seconds to time format (hours, minutes, and seconds)
seconds_to_time = function(seconds) {
  hours = as.integer(floor(seconds / 3600))
  minutes = as.integer(floor((seconds %% 3600) / 60))
  seconds = as.integer(seconds %% 60)
  sprintf("%02d:%02d:%02d", hours, minutes, seconds)}

sleepdf$WD_waketime_secs = time_to_seconds(sleepdf$WD_waketime)
mean_waketime_weekdays = mean(sleepdf$WD_waketime_secs)
CI_weekdays_waketime = t.test(sleepdf$WD_waketime_secs, conf.level = 0.95)$conf.int

seconds_to_time(mean_waketime_weekdays)
seconds_to_time(CI_weekdays_waketime)


#Mean waketime on weekends
sleepdf$WE_waketime_secs = time_to_seconds(sleepdf$WE_waketime)
mean_waketime_weekends = mean(sleepdf$WE_waketime_secs)
CI_weekends_waketime = t.test(sleepdf$WE_waketime_secs, conf.level = 0.95)$conf.int

seconds_to_time(mean_waketime_weekends)
seconds_to_time(CI_weekends_waketime)

#Difference between the waketime on weekdays and weekends
waketimedifference = seconds_to_time(mean_waketime_weekends-mean_waketime_weekdays)
waketimedifference

#MEAN SLEEP TIME
#Mean sleeptime on weekdays

sleepdf$WD_sleeptime_secs = time_to_seconds(sleepdf$WD_sleeptime)
mean_sleeptime_weekdays = mean(sleepdf$WD_sleeptime_secs)
CI_weekdays_sleeptime = t.test(sleepdf$WD_sleeptime_secs, conf.level = 0.95)$conf.int

seconds_to_time(mean_sleeptime_weekdays)
seconds_to_time(CI_weekdays_sleeptime)

#Mean sleeptime on weekends
sleepdf$WE_sleeptime_secs = time_to_seconds(sleepdf$WE_sleeptime)
mean_sleeptime_weekends = mean(sleepdf$WE_sleeptime_secs)
CI_weekends_sleeptime = t.test(sleepdf$WE_sleeptime_secs, conf.level = 0.95)$conf.int

seconds_to_time(mean_sleeptime_weekends)
seconds_to_time(CI_weekends_sleeptime)

#Difference between the waketime on weekdays and weekends
sleeptimedifference = seconds_to_time(mean_sleeptime_weekends-mean_sleeptime_weekdays)
sleeptimedifference

#Mean Social jetlag
socialjetlag_mean = mean(sleepdf$social_jetlag)
mean_social_jetlag_hours <- mean(sleepdf$social_jetlag) / 3600
socialjetlag_sd = sd(sleepdf$social_jetlag)
socialjetlag_n = length(sleepdf$social_jetlag)

# Calculating the standard error
socialjetlag_se = socialjetlag_sd / sqrt(socialjetlag_n)

# Calculating the 95% confidence interval using the t-distribution
alpha = 0.05
t_critical = qt(1 - alpha/2, df = socialjetlag_n - 1)
socialjetlag_conf_int = c(socialjetlag_mean - t_critical * socialjetlag_se, socialjetlag_mean +
                       t_critical * socialjetlag_se)

socialjetlag_mean
socialjetlag_conf_int


#Mean Sleep debt

mean_sleep_debt = mean(sleepdf$sleep_debt)
se_sleep_debt = sd(sleepdf$sleep_debt) / sqrt(nrow(sleepdf))

# Calculate the 95% confidence interval for sleep debt
CI_sleep_debt = c(mean_sleep_debt - qt(0.975, df = nrow(sleepdf) - 1) * se_sleep_debt,
                   mean_sleep_debt + qt(0.975, df = nrow(sleepdf) - 1) * se_sleep_debt)

mean_sleep_debt
CI_sleep_debt
```

```{r fig2, fig.width = 8, fig.asp = .62, echo=FALSE,  results='asis', fig.cap="Sleep debt compared to recommended 8 hours of sleep"}

# Define the recommended sleep duration (in hours)
recommended_sleep_duration = 8

# Calculate the absolute sleep debt for weekdays and weekends (in hours)
sleepdf = sleepdf %>%
  mutate(abs_sleep_debt_weekdays = ifelse(WD_sleephours / 60 < recommended_sleep_duration, recommended_sleep_duration - (WD_sleephours / 60), 0),
         abs_sleep_debt_weekends = ifelse(WE_sleephours / 60 < recommended_sleep_duration, recommended_sleep_duration - (WE_sleephours / 60), 0))

# Plot the absolute sleep debt for weekdays and weekends (in hours)
par(mfrow = c(1, 2))

hist(sleepdf$abs_sleep_debt_weekdays, main = "Absolute Sleep Debt on Weekdays", xlab = "Sleep Debt (hours)", col = "aquamarine3", breaks = 20)
hist(sleepdf$abs_sleep_debt_weekends,main = "Absolute Sleep Debt on Weekdays", xlab = "Sleep Debt (hours)", col = "aquamarine4", breaks = 20)
```

# 3 Statistical Methods
## Exploratory Data Analysis

Individuals who wake up at the same time on weekdays and weekends and get the same 
amount of sleep on both days are investigated to see if a regular sleep cycle reduces 
daytime sleepiness. To find the correlation, the ANOVA will compare the groups with 
and without a healthy sleep cycle. Given that the assumption of homogeneity of variance 
is not met for variance analysis, the Kruskal-Wallis test is utilized as a non-parametric alternative. The Kruskal-Wallis test indicates no significant difference in daytime 
`Oversleepy` between those with and without regular sleep cycles (p-value = 0.6854). 
Therefore, maintaining a regular sleep cycle does not significantly impact daytime 
sleepiness. As the chi-square is less for Kruskal-Wallis this forms one of the 
limitations of this statistical test. 

To determine the multiple linear regression, log transformation was done on `Snoring,` and multiple linear regression was done using variables like sleep debt, trouble sleeping, and over-sleepiness as multicollinearity was determined for them. The adjusted R square was 
determined by testing different models, and the adjusted r square for `WD_sleephours` 
was 18%, and for`WE_sleephours`, it was 30%, determining the best fit for the model. 
Outliers were also determined, but as the project deals with self-reported data, the 
outliers were not excluded from the dataset as they might also give important information. 
Robust regression was done to address the presence of outliers or influential points. 
For the robust regression, the residual error was less when compared to multiple linear regression,but the adjusted R square was low when compared to multiple linear models.
One of the limitations of this is the adjusted R square was less than the models. 
The normal Q-Q plots were plotted (See Figure 3), which slightly deviated from the 
normal, but as the population size is large, this was not considered.

For the logistic regression, a correlation heatmap was plotted to determine the highly 
associated factors. For the logistic regression, the binary variable `Trouble_sleep` 
was plotted against other independent variables, and the most association was found 
between `Oversleepy`, `Snoring`, and `Snorting`. The multiple logistic regression's 
independent variables were all categorical, so they were converted to factors to yield a 
best-fitting model. For this AIC value was 10726, which 
was best fitting when compared to other models.

## Limitations encountered during the project
1. As the data was self-reported, the outliers could not be removed, which might have significantly affected the data.
2. As the data was in time series, it was difficult to determine the exact sleep times 
as the meantime could not be exactly determined.
3. As the data were not normally distributed, ANOVA could not be tested to determine the 
mean across groups, and Kruskal-Wallis yielded a low chi-square value. 
4. The multiple linear regression yielded less adjusted R square for `WD_sleephours`, 
which might indicate certain missing data or other factors which are not associated with it.
5. The multiple logistic regression yielded high AIC due to the large sample size.



```{r fig3, fig.width = 7, fig.asp = .62, echo=FALSE,  results='asis', fig.cap="Normal Q-Q plots"}
# Convert Snoringto numeric (hours) and apply log transformation
sleepdf$log_Snoring <- log(sleepdf$Snoring + 1)
# For WD_sleephours model
wd_model <- lm(formula = WD_sleephours ~ log_Snoring + sleep_debt + Trouble_sleep + Oversleepy, data = sleepdf)

# For WE_sleephours model
we_model <- lm(formula = WE_sleephours ~ log_Snoring + sleep_debt + Trouble_sleep + Oversleepy, data = sleepdf)

# Custom Q-Q plot function
custom_qq_plot <- function(model, main_title, plot_color) {
  qqnorm(rstandard(model), main = main_title, col = plot_color, pch = 16)
  qqline(rstandard(model), col = "grey")
}

# Set up the plotting area with 1 row and 2 columns
par(mfrow = c(1, 2))

# Call the custom_qq_plot function to create the Q-Q plots for WD_sleephours and WE_sleephours
custom_qq_plot(wd_model, "WD_sleephours: Normal Q-Q", plot_color = "lightblue")
custom_qq_plot(we_model, "WE_sleephours: Normal Q-Q", plot_color = "navyblue")

# Reset the plotting area to the default single panel
par(mfrow = c(1, 1))

```

# 4 Results
In the US population from 2017 to 2021, the mean sleep duration on weekdays was 7.65 
hours (95% CI, 7.62-7.69 hours). The mean sleep duration on weekends was 8.36 hours 
(95% CI, 8.32-8.40 hours). The difference between weekday and weekend sleep duration was 0.70 hours. It indicated that individuals slept more on weekends than on weekdays. Sleep 
and wake times were also differed among the US population. The mean wake time on weekdays 
was 06:44:51 hours (95% CI, 06:42:10- 06:42:10 hours). The mean wake time on 
weekends was 07:52:34 hours (95% CI, 07:50:01-07:55:08 hours). The difference between  
wake times on weekdays and weekends was 01:07:43 hours.The mean sleep time on weekends 
and weekdays is 12:00 AM. The mean sleep debt was 0.50 hours (95% CI, 0.48-0.53 hours). 
The Kruskal-Wallis test was performed to determine whether individuals with regular sleep 
cycles also experience excessive daytime sleepiness. The p-value of 0.6851 indicates that 
there was no statistically significant difference in the distribution of Oversleepy scores 
between those with regular sleep cycles and those without, so failing to reject the null hypothesis where regular sleep cycle is [regular_sleep_cycle = (Weekday_waketime == Weekend_waketime) & (Weekday_sleephours == Weekend_sleephours)].

The multiple linear regression was done on between the dependent variable of weekday
sleep hours and several independent variables, including log-transformed snoring, 
sleep debt, trouble sleeping, and over-sleepiness. The statistical analysis reveals 
that the model as a whole is statistically significant (F-statistic = 538.3, p < 2.2e-16). Moreover, the adjusted R-squared value suggests that the explanatory variables account 
for approximately 18% of the variability in the response variable. The regression 
analysis reveals that the variable of snoring, which has been subjected to log transformation, exhibits a negative coefficient of -0.17192. This suggests an inverse relationship 
exists between the level of snoring and the number of hours of sleep during weekdays, 
whereby an increase in snoring is associated with a decrease in weekday sleep hours. 
The findings reveal a negative correlation between sleep debt and weekday sleep hours, as evidenced by the negative coefficient value of -0.57194. This suggests that an increase 
in sleep debt is associated with a decrease in the number of hours slept during weekdays. 
The coefficient for trouble_sleep is positively significant (0.15174), implying that an 
increase in the difficulty level is associated with an increase in the number of hours
of sleep during weekdays. The variable Oversleepy exhibits a negative coefficient value of -0.05041, which suggests an inverse relationship between the degree of over sleepiness
and the duration of sleep hours on weekdays. Another multiple linear regression was 
done on Weekend sleephours and log-transformed snoring, sleep debt, trouble sleeping, 
and over-sleepiness which resulted in Multiple R-squared:0.3071, Adjusted R-squared: 0.3068 indicating a best fitting model when compared to others.

After performing the correlation for the binary variable `Trouble_sleep` to other 
variables like `Oversleepy`, `Snoring`, `Snorting`, `sleep_debt`, `WD_sleephours` and `WE_sleephours`. The correlation was found for `Trouble_sleep`, `Oversleepy`, `Snoring`, `Snorting` and multiple logistic regression was performed which yielded. The findings 
indicate that with the exception of Oversleepy9 and Snoring1, all predictor variables 
demonstrated statistical significance in predicting trouble_sleep. The study reported 
the odds ratio for each level of the predictor variables along with their corresponding 
p-values. The adequacy of the model fit was evaluated by means of the deviance residuals 
and the AIC value which is 10726. The adequacy of the model was demonstrated by the
residual deviance being less than the null deviance, and the AIC value being comparatively 
low when compared to other models.The results indicate that trouble sleeping is associated
with snoring, snorting, and excessive sleepiness during the day.

# 5 Strengths and Limitations of the project
One of the project's strengths is that it is based on a significant and diverse 
population. There are, however, several limitations to consider. 
Firstly, the study depends on self-reported sleep data, which is prone to memory 
bias because individuals may not remember their sleep periods precisely, resulting 
in misinterpretation of results. Second, missing data were eliminated from the study, 
which may have resulted in the loss of crucial information. However, keeping it may 
not have improved results in this context.

   Furthermore, the study may not have considered all confounding factors that could 
influence sleep deprivation, such as profession, age, gender, and multiple disorders. 
Including these characteristics could provide a complete knowledge of people's sleeping
habits. Finally, because the data format for sleep timings (hh: mm) may not accurately 
reflect the sleep duration or the precise sleep and wake times, it may have contributed
to the misinterpretation of the results.

  Despite these limitations, the study provides valuable information about sleep patterns
and factors influencing sleep quality. Future studies could overcome these limitations
by including more diverse factors, upgrading data collection methods, and analyzing 
the data using better statistical methods.

# 6 Conclusion
The project found significant disparities in sleep habits between weekdays and 
weekends among the US population. The results indicated that individuals tend to 
sleep longer on weekends than during the week, which is consistent with the notion 
of "sleep debt." Another important finding was the presence of social jet lag, 
which suggests that social jet lag is a prevalent phenomenon among the US population, 
with potentially detrimental effects on health and well-being. This research has 
significant implications for comprehending sleep patterns and their effects on health. 
Healthcare professionals can devise targeted interventions to promote healthy sleep 
habits and reduce the risk of sleep-related disorders by identifying weekday and 
weekend sleep differences. In addition, this project's findings can be used to 
investigate other factors that may contribute to sleep deprivation, such as shift 
work and stress, and to develop individualized strategies to improve sleep patterns 
among the US population.



















# References
Di, H., Guo, Y., Daghlas, I., Wang, L., Liu, G., Pan, A., Liu, L., & Shan, Z. (2022). Evaluation of sleep habits and disturbances among US adults, 2017-2020. JAMA network open, 5(11), e2240788-e2240788. https://doi.org/10.1001/jamanetworkopen.2022.40788 \

Ford, E. S., Cunningham, T. J., & Croft, J. B. (2015). Trends in self-reported sleep duration among US adults from 1985 to 2012. Sleep, 38(5), 829-832. https://doi.org/10.5665/sleep.4684 \

Grandner, M. A. (2017). Sleep, health, and society. Sleep medicine clinics, 12(1), 1-22. https://doi.org/10.1016/j.jsmc.2016.10.012 \

Roenneberg, T., Allebrandt, Karla V., Merrow, M., & Vetter, C. (2012). Social jetlag and obesity. Current Biology, 22(10), 939-943. https://doi.org/10.1016/j.cub.2012.03.038 



# Appendix
```{r fig.width = 5, fig.asp = .62,echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results='asis'}
##Plotting sleep duration graphs

# Creating the bar graph
#Loading the required library
library(ggplot2)
# Creating a data frame for the means
mean_sleep_data = data.frame(
  Sleep = c("Weekdays", "Weekends"),
  MeanSleepDuration = c(weekday_mean, weekend_mean))

ggplot(mean_sleep_data, aes(x = Sleep, y = MeanSleepDuration)) +
  geom_bar(stat = "identity", fill = "darkseagreen3") +
  geom_text(aes(label = round(MeanSleepDuration, 2)), vjust = -0.5, size = 4) +
  ylim(0, max(mean_sleep_data$MeanSleepDuration) * 1.2) +
  labs(title = "Mean Sleep Duration", x = "Sleep", y = "Mean Sleep Duration (hours)") +
  theme_minimal()
```

```{r fig.width = 5, fig.asp = .62, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results='asis'}
# Load the ggplot2 package
library(ggplot2)

# Create a temporary variable for the purpose of the plot
sleepdf_temp <- sleepdf %>%
  mutate(Snoring_labels = factor(Snoring, levels = c(0, 1, 2, 3, 7, 9),
                                 labels = c("Never", "Rarely", "Occasionally", "Frequently", "Refused", "Don't know")))

# Create a bar plot using ggplot2
ggplot(sleepdf_temp, aes(x = Snoring_labels)) +
  geom_bar(fill = "rosybrown1") +
  labs(title = "Snoring Frequency",
       x = "Snoring Scale",
       y = "Frequency") +
  theme_minimal()
```

```{r fig.width = 6, fig.asp = .75, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results='asis'}
# Create a temporary variable for the purpose of the plot
sleepdf_temp_snorting <- sleepdf %>%
  mutate(Snorting_labels = factor(Snorting, levels = c(0, 1, 2, 3, 7, 9),
                                 labels = c("Never", "Rarely", "Occasionally", "Frequently", "Refused", "Don't know")))

# Create a bar plot using ggplot2
ggplot(sleepdf_temp_snorting, aes(x = Snorting_labels)) +
  geom_bar(fill = "lightblue") +
  labs(title = "Snorting Frequency",
       x = "Snorting Scale",
       y = "Frequency") +
  theme_minimal()

# Create a temporary variable for the purpose of the plot
sleepdf_temp_oversleepy <- sleepdf %>%
  mutate(Oversleepy_labels = factor(Oversleepy, levels = c(0, 1, 2, 3, 4, 7, 9),
                                  labels = c("Never", "Rarely", "Sometimes", "Often", "Almost always", "Refused", "Don't know")))

# Create a bar plot using ggplot2
ggplot(sleepdf_temp_oversleepy, aes(x = Oversleepy_labels)) +
  geom_bar(fill = "palegreen2") +
  labs(title = "Oversleepy Frequency",
       x = "Oversleepy Scale",
       y = "Frequency") +
  theme_minimal()

#For trouble sleep
count_table <- table(sleepdf$Trouble_sleep)

# Create a bar plot for the Trouble_sleep variable
barplot(count_table, main = "Trouble Sleeping Distribution", xlab = "Category", ylab = "Frequency", names.arg = c("Yes", "No", "Refused", "Don't know"), col = "plum")
```

```{r echo=FALSE,  message=FALSE, warning=FALSE, paged.print=FALSE}
# Identifying individuals with regular sleep cycles
sleepdf = sleepdf %>%
  mutate(regular_sleep_cycle = (WD_waketime == WE_waketime) & (WD_sleephours == WE_sleephours))

# Kruskal-Wallis test
kruskal_wallis_result <- kruskal.test(Oversleepy ~ regular_sleep_cycle, data = sleepdf)
cat("kruskal_wallis test for regular sleep cycle and oversleepy:\n\n")
kruskal_wallis_result
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Convert Snoringto numeric (hours) and apply log transformation
sleepdf$log_Snoring <- log(sleepdf$Snoring + 1)


## Fit a new multiple linear regression model with transformed variables
log_transformed_model_WD <- lm(WD_sleephours ~ log_Snoring  + sleep_debt + Trouble_sleep+ Oversleepy, data = sleepdf)
cat("Multiple Linear Regression for weekdays:\n\n")
summary(log_transformed_model_WD)

```

```{r, echo= FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
log_transformed_model_WE <- lm(WE_sleephours ~ log_Snoring  + sleep_debt + Trouble_sleep+ Oversleepy, data = sleepdf)
cat("Multiple Linear Regression for weekends:\n\n")
summary(log_transformed_model_WE)
```

```{r, echo= FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
# Residuals vs Fitted
par(mfrow = c(2, 2))
custom_residuals_vs_fitted <- function(model, main_title) {
  plot(model$fitted.values, rstandard(model),
       main = main_title,
       xlab = "Fitted Values",
       ylab = "Standardized Residuals",
       col = "lightblue",
       pch = 16)
  abline(h = 0, lty = 2, col = "gray") # Add a dashed line at y = 0
}

# Call the custom_residuals_vs_fitted function to create the plot for WD_sleephours
custom_residuals_vs_fitted(log_transformed_model_WD, "WD_sleephours: Residuals vs Fitted")

# Call the custom_residuals_vs_fitted function to create the plot for WD_sleephours
custom_residuals_vs_fitted(log_transformed_model_WE, "WD_sleephours: Residuals vs Fitted")
```

```{r echo=FALSE,  results='asis', fig.cap="Correlation plot"}


# Create a new binary variable based on the Trouble_sleep variable
sleepdf$binary_trouble_sleep <- ifelse(sleepdf$Trouble_sleep == 1, 1, 0)

# Remove rows with "refused" (7) or "don't know" (9)
clean_sleepdf <- sleepdf[sleepdf$Trouble_sleep %in% c(1, 2), ]


# Create a correlation matrix
cor_matrix <- cor(clean_sleepdf[, c("binary_trouble_sleep", "Oversleepy", "Snoring", "Snorting", "sleep_debt", "WD_sleephours", "WE_sleephours")])
```



```{r echo= FALSE, message=FALSE, warning=FALSE}
# Convert Oversleepy and Snoring to factors
clean_sleepdf$Oversleepy <- as.factor(clean_sleepdf$Oversleepy)
clean_sleepdf$Snoring <- as.factor(clean_sleepdf$Snoring)
clean_sleepdf$Snorting <- as.factor(clean_sleepdf$Snorting)

# Run the logistic regression model
logistic_model_factors <- glm(binary_trouble_sleep ~ Oversleepy + Snoring + Snorting, family = binomial, data = clean_sleepdf)

# Display the summary of the model
cat("Multiple Logistic Regression:\n\n")
summary(logistic_model_factors) 
```